---
title: "glm Cheat Sheet"
author: "JCSalo"
date: "August 1, 2017"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(survival)
library(broom)
```

## R Markdown

The	glm	function	is	the	basic	tool	for	fitting	generalized	linear	models:

> glm(formula, family, data, weights, subset, ...)

The	family	parameter	is	specific	to	the	glm	function.	There are	six	choices	of	family:	

+  family = gaussian (link=identity)
+  family = binomial (link=logit, probit, or cloglog)
+  family = poisson (link=log, identity or sqrt)
+  family = Gamma (link = inverse, identity, or log)
+  family = inverse.gaussian (link=1/mu^2)

The	family	parameter	is	specific	to	the	glm	function.	There are	six	choices	of	family:	
Each	of	the	first	five	choices	has	a	variance	function	and	one	or	more	link	functions.	The	
last	is	user-defined.	To	use	default	link,	all	you	have	to	do	is	specify	the	family	name.	To	use	
an	alternative	link	you	must	specify	what	you	want:


```{r glm_meal, comment=''}
glm_meal<-glm(ph.ecog ~ meal.cal+ wt.loss + inst, data=cancer)
glm_meal
summary(glm_meal) # displays detailed model information
summary(glm_meal)$coefficients # shows coefficients in a matrix
summary(glm_meal)$coefficients[-1,c('Estimate','Pr(>|t|)')] # selects cells from matrix
tidy(glm_meal) # 'tidy' display of results
```

## Univariate glm - interative loop

You can also embed plots, for example:

```{r uniglmloop, comment=''}

ph.ecog_fit<-""  #create a variable (dataframe) which will hold the glm data

for (vars in c("meal.cal","wt.loss","age","inst","sex" )){  #for loop iterates through variables
  
  ph.ecog_uniglm<-glm(ph.ecog ~ get(vars), data=cancer) # call glm function. 
      # Note "get(vars)" used for the independent variable
  ph.ecog_uniglm_tidy<-tidy(ph.ecog_uniglm)  # tidy() function creates a compact output of results
  ph.ecog_uniglm_tidy<-ph.ecog_uniglm_tidy[-1,]  # delete the first line of glm() output 
      # (which contains the intercept)
  ph.ecog_uniglm_tidy[c(2:5)]<-lapply(ph.ecog_uniglm_tidy[c(2:5)], function(x) signif(x,digits=5) )  
      #limit data to 5 significant digits
  col1<-paste(vars,substr(ph.ecog_uniglm_tidy[,1],10,100),sep="|") # creates a label for the output line
      # IF variables used are factors (see below), we can shows them
  ph.ecog_uniglm_tidy[,1]<-col1   # moves the label into the first column of the glm() output
  ph.ecog_fit<-rbind(ph.ecog_fit,ph.ecog_uniglm_tidy)  # adds the output line to the dataframe of results
 
}
print(ph.ecog_fit, row.names=FALSE) # print the dataframe, omiting the row numbers
```

If variables for glm as represented as factors, glm() will show the results for each level of the factor:

We will create a new variable sexf which is a factor with two levels

We will also change wt.loss into a categorical variable with cut()

```{r uniglmloop3, comment=''}

cancer$sexf<-factor(cancer$sex, levels = c(1,2), labels = c("M","F"))
cancer$wt.loss_cut<-cut(cancer$wt.loss, breaks=quantile(cancer$wt.loss
    , probs=seq(0,1,by=0.25),na.rm=TRUE)
    ,labels=c('Low','Med','High','vHigh'),include.lowest=TRUE)
 table(cancer$wt.loss_cut,dnn=c("wt.loss_cut"))
```
Variable sex is of class `r class(cancer$sex)`

Variable sexf is of class `r class(cancer$sexf)`

Variable wt.loss is of class `r class(cancer$wt.loss)`

Variable wt.loss_cut is of class `r class(cancer$wt.loss_cut)`

```{r uniglmloop3a, comment=''}
ph.ecog_fit<-""  

for (vars in c("meal.cal","wt.loss_cut","age","inst","sexf" )){  #for loop iterates through variables
  
  ph.ecog_uniglm<-glm(ph.ecog ~ get(vars), data=cancer)
  ph.ecog_uniglm_tidy<-tidy(ph.ecog_uniglm)  
  ph.ecog_uniglm_tidy<-ph.ecog_uniglm_tidy[-1,] 
  ph.ecog_uniglm_tidy[c(2:5)]<-lapply(ph.ecog_uniglm_tidy[c(2:5)], function(x) signif(x,digits=5) )  
  col1<-paste(vars,substr(ph.ecog_uniglm_tidy[,1],10,100),sep="|") # creates a label for the output line
    # We can now see the factor levels displayed in results
  ph.ecog_uniglm_tidy[,1]<-col1   
  ph.ecog_fit<-rbind(ph.ecog_fit,ph.ecog_uniglm_tidy) 
}
print(ph.ecog_fit, row.names=FALSE) # print the dataframe, omiting the row numbers
```

Note that 'Low' is chosen as the referent, because it is the first label

We can reorder the factor as follows:

```{r uniglmloop4, comment=''}
cancer$wt.loss_cutR<-relevel(cancer$wt.loss_cut, "vHigh")
ph.ecog_fit<-""  

for (vars in c("meal.cal","wt.loss_cutR","age","inst","sexf" )){  #for loop iterates through variables
  
  ph.ecog_uniglm<-glm(ph.ecog ~ get(vars), data=cancer) 
  ph.ecog_uniglm_tidy<-tidy(ph.ecog_uniglm)  
  ph.ecog_uniglm_tidy<-ph.ecog_uniglm_tidy[-1,] 
  ph.ecog_uniglm_tidy[c(2:5)]<-lapply(ph.ecog_uniglm_tidy[c(2:5)], function(x) signif(x,digits=5) )  
  col1<-paste(vars,substr(ph.ecog_uniglm_tidy[,1],10,100),sep="|") 
    # Note the change in levels displayed for wt.loss_cutR
  ph.ecog_uniglm_tidy[,1]<-col1   
  ph.ecog_fit<-rbind(ph.ecog_fit,ph.ecog_uniglm_tidy) 

}
print(ph.ecog_fit, row.names=FALSE) # print the dataframe, omiting the row numbers
```

What we see here is that globally, wt.loss is a significant predictor of ECOG performance status.  (p=0.006). 

When Low weight loss is used as the referent, Med and High are not different, but vHigh is different.

When vHigh weight loss is used as a referent, it is clear that it is different from Low, Med, and High
