---
title: "CheatSheet"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse) #Tools for tidy data analysis - Hadley Wickham
library(survival)
library(knitr) # Used to generate reproduceable rearch
library(pander) # Generate tables for MS Word documents
library(tableone) # Generates Table 1 for clinical research reporting
library(descr)  # Used for CrossTable
library(broom)  # used for tidy() - function which reformats regression coefficients 

```

# Introduction

## Getting Help
?mean  Get help in a particular function

## File Navigation

getwd() | Show working directory

setwd('H://file/path') | Set working directory

R Studio: Create a project by pointing to an existing directory of scripts

# Core Programming

## Variables
```{r, echo=TRUE}
x<-2L
typeof(x)
class(x)
str(x)
y<-2.5
typeof(y)
z<-"h"
typeof(z)
a<-T
typeof(a)
b<-FALSE
typeof(b)

```

## Using Variables
```{r, echo=TRUE}
greeting<- "hello"
recipient <- "World"
message<-paste(greeting, recipient)
message
```

## Logical Operators

Pipe is OR logical operator

Ampersand is AND logical operator

```{r, echo=TRUE}
result<- 4<5
result

result2 <- !(5>1)
result2

result | result2

result & result2
```

# While and For Loops

Note 1:5 is a vector (one-dimensional array)

```{r}

counter <-1 
while (counter <12) {
  print (counter)
  counter<- counter +1
}

for (i in 1:5) {
  print("Hellow World")
}

1:5
```


```{r}


x<-rnorm(1)
if (x>1) {
  answer <- "Greater than 1"
}else if (x>=-1){
  answer <- "Between -1 and 1"
} else {
  answer <-"Less than -1"
}
x
answer

```

## Vectors

Note bracket indexing - Different lists/vectors will behave differently with these two notationss
```{r}
vector1<- c(3,4,56, 72,125, 2)
vector1
typeof(vector1)
str(vector1)
class(vector1)
is.numeric(vector1)
is.integer(vector1)
is.character(vector1)


vector2<- c("G",4,"Smith", 72)
vector2
typeof(vector2)
str(vector2)
class(vector2)
is.numeric(vector2)
is.integer(vector2)
is.character(vector2)

vector3<-seq(1,9,by=2)
vector3

vector1[2]
vector1[[2]]

vector1[-3]

vector1[2:4]

vector1[c(1,3,4)]
```
## Functions

```{r}

A<- c(1,3,5,7,9,11)
B<-sqrt(A)
B
```

## Packages
```{r}

try(?ggplot2())
library(ggplot2)
?ggplot2()
qplot(data=diamonds, carat, price, colour=clarity, facets=.~clarity)

```

## Load built-in dataset and explore
```{r load, warning=FALSE}
data(cancer) # load the dataset 'cancer' (from the 'survival' package)
colnames(cancer) # snows names of colums
summary(cancer$age) # shows summary of the contents of each column
median(cancer$age,na.rm=TRUE) # Need to remove any NA values
str(cancer) # shows the structure of the object (dataframe).


```

# Tidyverse

Create tibble from dataframe

```{r tibble}
cancert<-as_tibble(cancer)
cancert
```



```{r save}
#library(openxlsx)
#library(xlsx)
#library(xlsxjars)
#library(WriteXLS)
#WriteXLS(cancer, ExcelFileName="cheat.xls", perl="perl", verbose=TRUE )
#write.xlsx(cancer,file="cheat.xlsx", col.names=TRUE, row.names=TRUE, append=FALSE, showNA=FALSE)

```


## Factors, Cut, Recode
```{r factors, echo=TRUE, comment=NA}
#create factors
cancer$sex_F<-factor(cancer$sex, levels = c(1,2), labels=c("Male","Female"))
class(cancer$sex)
class(cancer$sex_F)

#Cut to transform continuous variable to categorical
cancer$agegroup_R<-cut(cancer$age, c(0,25,50,75,110))

#Recoding using ifelse
cancer$age_over_65 <- ifelse (cancer$age %in% c(65:100), "Over65", "Under65")

#Recoding using bracket notation
cancer$age_over_75 <- 0
cancer$age_over_75[cancer$age >75]<-1
```

## Histogram
```{r histogram, echo=TRUE, comment=NA}
hist(cancer$age)
hist(cancer$age,freq=FALSE, xlab = "Age", main="Distribution of Age",sub="Cancer Dataframe", col="light blue", xlim=c(2,100), ylim=c(0,0.05))
curve(dnorm(x, mean=mean(cancer$age), sd = sd(cancer$age)), add=TRUE, col="darkblue", lwd=2)
```

## Scatterplot

```

## Scatter Plot - Data Subsets

```

## Scatter Plots


```{r cars}
#data(cancer)
colnames(cancer)
table(cancer$meal.cal)
plot(cancer$meal.cal ~ cancer$age, xlab="Age", ylab="Meal Calories", ylim=c(0,2600), xlim=c(0,100), main="Meal Calories ~ Age")
abline(lm(cancer$meal.cal ~ cancer$age), lty=2)
```

## Overlay Scatter Plots

```{r scatteroverlay, echo=FALSE}
plot(cancer$meal.cal[cancer$sex==1] ~ cancer$age[cancer$sex==1], xlab="Age", ylab="Meal Calories", ylim=c(0,2600), xlim=c(0,100),col="blue")
abline(lm(cancer$meal.cal ~ cancer$age), lty=2)
par(new=TRUE)
plot(cancer$meal.cal[cancer$sex==2] ~ cancer$age[cancer$sex==2], xlab="", ylab="", ylim=c(0,2600), xlim=c(0,100),col="red",axes=FALSE)



```





## Tables
```{r pressure, echo=TRUE, comment=NA}
table(cancer$sex_F)
addmargins(table(cancer$sex_F))
table(cancer$sex_F,cancer$ph.ecog,dnn=c("Gender","ECOG PS"))
#
```

## Subsets - using subset()
(Note that because sex_F is a factor, all levels are shown in the table)
```{r subsets, comment = NA}
cancer.mecog1 <-subset(cancer,ph.ecog==1 & sex_F =="Male" ) 
table(cancer.mecog1$sex_F,cancer.mecog1$ph.ecog,dnn=c("Gender","ECOG PS"))
```


## Subsets - alternate
(Note that because sex_F is a factor, all levels are shown in the table)
```{r subsetalt, comment=NA}
table(cancer$sex_F,cancer$ph.karno,dnn=c("Gender","Physician Karnofsky Scale"))
cancer.karno7090 <-cancer[cancer$ph.karno %in% c(70,80,90),]
table(cancer.karno7090$sex_F,cancer.karno7090$ph.karno,dnn=c("Gender","Physician Karnofsky Scale"))
```



## Slice dataframe into vectors

Bracket notation slices a dataframe [rows,colums]

cancer[,2]  selects all rows and only column 2
cancer[,"sex"]  selects all rows and only the colum labeled "sex"
cancer[3,]  selects all columns of the third row of the dataframe
cancer[cancer$sex==1,] selects all columns where all rows satisfy cancer$sex==1
```{r slice, echo=TRUE, comment = NA}
cancer.femalekarno70 <-cancer[cancer$ph.karno %in% c(70) & cancer$sex ==2,]
print (cancer.femalekarno70[,c("sex","ph.ecog")], row.names=FALSE)
# Supress printing of row names
```


## Comparision of categorical variables between groups
### Chi Square and Fisher
```{r chisq,echo=TRUE, comment=NA}
chisq.test(table(cancer$sex_F,cancer$ph.ecog))
chisq.test(table(cancer$sex_F,cancer$ph.ecog))$p.value
fisher.test(table(cancer$sex_F,cancer$ph.ecog))
fisher.test(table(cancer$sex_F,cancer$ph.ecog))$p.value
```

## Comparison of continuous variables between groups
### Wilcox and Kruskal tests
```{r wilcox, echo=TRUE, comment=NA}
wilcox.test(cancer$meal.cal ~ cancer$sex_F)
kruskal.test(cancer$meal.cal ~ cancer$sex_F)
boxplot(cancer$meal.cal ~ cancer$sex_F, xlab="Gender", ylab="Meal Calories")
```


### CrossTable (requires descr package)
Parameters allow precise control of display of row vs column percentages (and totals)
```{r, echo=TRUE, comment=NA}
CrossTable(cancer$sex_F,cancer$ph.ecog)

ct<-CrossTable(cancer$sex_F,cancer$ph.ecog, digits = list(expected=1, prop=3,percent=2,others=3),prop.r=TRUE,prop.c=FALSE,prop.t=FALSE, prop.chisq=FALSE, chisq=TRUE, missing.include=FALSE, dnn=c("Gender", "ECOG"),cell.layout=FALSE,total.r=TRUE,total.c=TRUE)
plot(ct, inv.y=TRUE)
print(ct)

```

```{r, include=FALSE}
# To extract coefficients from fitted model

#coxout<-coxph(Surv(t,t.censor)~x)

#coxout$coef
#coxout$p
#summary(coxout)$logtest["test"]
#summary(coxout)$logtest["pvalue"]
#Can learn more:
#str(summary(coxout))

```


### Pander output CrossTable
```{r , echo=FALSE,results='asis', warning=FALSE, message=FALSE}
#panderOptions('digits',2)
#panderOptions('round',2)
#panderOptions('keep.trailing.zeros',TRUE)
#pander(ct,caption=attr(ct,"CrossTable of Gender vs ECOG"),total.r=ct$total.r,total.c=ct$total.c)
```



# Pander Output
### Pander output of Chi square
```{r , echo=FALSE,results='asis', warning=FALSE}
#pander(chisq.test(table(cancer$sex_F,cancer$ph.ecog)),caption="Chi Square Test Sex ~ ECOG")
```


### Pander output of simple tables
```{r , echo=FALSE,results='asis'}
#pander(table(cancer$sex_F,cancer$ph.ecog,dnn=c("Gender","ECOG PS")),caption="Table of Sex x ECOG", dnn=c("Sex", "ECOG"))
```


```{r foreignout, echo=FALSE}
#library(foreign)
#write.foreign(rectum,datafile= "../derived/rectum9814RSD.txt",codefile =  "../derived/rectum9814RSD.sas", package = "SAS")
```


## Tableone - Stratified
```{r tableone, echo=FALSE, comment=NA, warning=FALSE, results='asis'}
myVars <- c("time", "status", "trt", "age", "sex", "ascites", "hepato",
          "spiders", "edema", "bili", "chol", "albumin", "copper", "alk.phos",
          "ast", "trig", "platelet", "protime", "stage")
catVars <- c("status", "trt", "ascites", "hepato",
             "spiders", "edema", "stage")
biomarkers <- c("bili","chol","copper","alk.phos","ast","trig","protime")
tab3 <- CreateTableOne(vars = myVars, strata = "trt" , data = pbc, factorVars = catVars)
printtab3 <- print(tab3,printToggle=FALSE,exact="stage")
pander(printtab3)
```


## Tableone - Include non-normal continuous varaibles
```{r tableone4, echo=FALSE, comment=NA, warning=FALSE, results='asis'}
myVars <- c("time", "status", "trt", "age", "sex", "ascites", "hepato",
          "spiders", "edema", "bili", "chol", "albumin", "copper", "alk.phos",
          "ast", "trig", "platelet", "protime", "stage")
catVars <- c("status", "trt", "ascites", "hepato",
             "spiders", "edema", "stage")
biomarkers <- c("bili","chol","copper","alk.phos","ast","trig","protime")
tab3 <- CreateTableOne(vars = myVars , data = pbc, factorVars = catVars)
printtab3 <- print(tab3,printToggle=FALSE,nonnormal=biomarkers,exact="stage")
pander(printtab3)
```



##Tableone - Categorical Variables Only

```{r tableone3,echo=FALSE}
printtab3cat <- print(tab3$CatTable,printToggle=FALSE)
pander(printtab3cat)

```

## Tableone - Continuous variables only
```{r tableone5,echo=FALSE}
printtab4cont <- print(tab3$ContTable,printToggle=FALSE)
pander(printtab4cont)

```



```{r SASexportout, echo=FALSE}
#library(SASexport)
#write.export(rectum, file= "../derived/rectum9814RSD.xpt")

```
